%\documentclass[12pt]{aastex}
\documentclass[iop, numberedappendix]{emulateapj}
%\documentclass{article}
%\usepackage{draftwatermark}
%\usepackage{paralist}
\shorttitle{Isothermal dusty gas dynamics}
\shortauthors{M-K.\ Lin, A.N. Youdin}

\usepackage{amsmath}
\usepackage{paralist}
\usepackage{url}
\usepackage{bm}
%\usepackage{hyperref}
%\usepackage[font=small,format=plain,labelfont=bf,up,textfont=it,up]{caption}
\newcommand{\p}{\partial}
\newcommand{\zmax}{z_\mathrm{max}}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\sbar}{\bar{\sigma}} 
\newcommand{\He}{\operatorname{He}}
\newcommand{\imgi}{\mathrm{i}} 
\newcommand{\ii}{\mathrm{i}}
\newcommand{\C}{\mathcal{C}^\lambda}
\newcommand{\wop}[1]{\mathcal{V}^{(#1)}}
\newcommand{\avg}[1]{\langle{#1}\rangle}
%\newcommand{\max}[1]{\mathrm{max}({#1})}
%\newcommand{\min}[1]{\mathrm{min}({#1})}
\newcommand{\zhat}{\hat{z}}
\newcommand{\khat}{\hat{k}}
\newcommand{\rhat}{\hat{r}}
\newcommand{\he}{\mathcal{H}}
\newcommand{\amp}{\mathcal{A}}
\newcommand{\dd}{\delta}
\newcommand{\real}{\operatorname{Re}}
\newcommand{\imag}{\operatorname{Im}}
\newcommand{\sgn}{\operatorname{sgn}}
\newcommand{\wbar}{\tilde{W}}
\newcommand{\qbar}{\tilde{Q}}
\newcommand{\dbigH}{\frac{H^\prime}{H}}
\newcommand{\Rp}{\frac{\rho_0^\prime}{\rho_0}}
\newcommand{\Fp}{\frac{g^\prime}{g}}
\newcommand{\Rpp}{\frac{\rho_0^{\prime\prime}}{\rho_0}}
\newcommand{\Fpp}{\frac{g^{\prime\prime}}{g}}
\newcommand{\ils}{L_s^{-1}}
\newcommand{\ilp}{L_p^{-1}}
\newcommand{\ihp}{H_p^{-1}}
\newcommand{\ihs}{H_s^{-1}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\csmid}{c_{s0}} 
\newcommand{\hiso}{h_\mathrm{iso}}
\newcommand{\Hiso}{H_\mathrm{iso}}
\newcommand{\zeus}{\texttt{ZEUS-MP }}
\newcommand{\rsph}{r_\mathrm{sph}}
\newcommand{\dvx}{\dd v_x}
\newcommand{\dvy}{\dd v_y}
\newcommand{\dvz}{\dd v_z}
\newcommand{\dbx}{\dd B_x}
\newcommand{\dby}{\dd B_y}
\newcommand{\dbz}{\dd B_z}
\newcommand{\Htilde}{H_\epsilon}
\newcommand{\Hdust}{H_\mathrm{d}}
\newcommand{\Hgas}{H_\mathrm{g}}
\newcommand{\hgas}{h_\mathrm{g}}
\newcommand{\tepsilon}{\tilde{\epsilon}}
\newcommand{\rhod}{\rho_\mathrm{d}}
\newcommand{\rhog}{\rho_\mathrm{g}}
\newcommand{\tstop}{t_\mathrm{s}}
\newcommand{\w}{ \widetilde{W}}
\newcommand{\dphi}{\dd \Phi}
\newcommand{\Hg}{H_\mathrm{g}}
\newcommand{\Hd}{H_\mathrm{d}} 
\newcommand{\actaa}{Acta Astron.}
\defcitealias{lin12}{L12} 
\defcitealias{lin15}{LY15}
\defcitealias{nelson13}{N13}
\defcitealias{barker15}{BL15}
\defcitealias{mcnally14}{MP14}
\defcitealias{goldreich67}{GS67}

\def \OmK {\Omega_{\rm K}}

\begin{document}

%\title{On the entropy of 
%  isothermal dusty gas: application to protoplanetary disks}  
\title{A thermodynamic view of dusty protoplanetary disks:
  axisymmetric stability}  
\author{Min-Kai Lin\altaffilmark{1} and Andrew N. Youdin}
\affil{Department of Astronomy and Steward Observatory, \\
University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\altaffiltext{1}{Steward Theory Fellow}
\email{minkailin@email.arizona.edu}

\begin{abstract}
  Small solids embedded in gaseous protoplanetary disks are subject
  to strong dust-gas friction. Consequently, tightly-coupled dust
  particles almost follows the gas flow. 
  This is analogous to the 
  conservation of entropy 
  following an adiabatic fluid.  We show that in the strong drag
  limit that an isothermal dusty gas indeed behaves like an  
  adiabatic pure gas, and that finite dust-gas coupling may be
  regarded as an effective 
  heating/cooling. We exploit this equivalence 
  to 
%\begin{inparaenum}[1)]
%\item
%  identify the entropy of isothermal dusty gas; and 
%\item 
  generalize previous, well-known gaseous instabilities
  to dusty protoplanetary disks.
%\end{inparaenum}
%This includes the axisymmetric stability of strictly isothermal dusty
%gas, allowing us to constrain the sharpness of 
%dust edges that can persist in protoplanetary disks. 
We show that strictly isothermal, non-self-gravitating, axisymmetric
disks perfectly coupled to dust are generally stable, however thin the
dust layer is.  %except possibily in the presence of sharp 
%edges in radial dependence of the dust-to-gas ratio. 
We also develop 
the dusty vertical shear instability (VSI) in locally isothermal
disks. We find that  dust-loading reduces VSI growth rates, and
preferentially stabilizes VSI modes with small radial 
wavelengths, as well as `surface modes' concentrated near the
disk boundaries. These results can be understood in terms of 
the buoyancy frequency of a isothermal dusty disk. 
%We show that vertical shear arising from
%dust settling does not lead to axisymmetric instability. 
We briefly discuss potential thermodynamical interpretations of 
dust-gas drag dynamics, such as the streaming instability in
protoplanetary disks as a manifestation of pulsational instabilities
in stars.  
\end{abstract}

\section{Introduction}
%si, sgi as examples
Protoplanetary disks are comprised of a mixture of gas and dust
\citep{chiang10}. Although the overall dust content is small
($\sim 1\%$ by mass), their presence can have profound impacts on the 
gas dynamics and vice versa. Dust-gas drag introduces a 
wide range of new phenomena that are absent in pure gas
disks. Important examples include the streaming instability
\citep{youdin05a,youdin07b,johansen07}, secular gravitational
instabilities \citep{ward00,youdin11,michikoshi12,takahashi14} and
Kelvin-Helmholtz instabilities
\citep{chiang08,barranco09,lee10}. These lead to structure and
planetesimal formation, as well as turbulence. More recently, new
dusty instabilities have also been uncovered
\citep{loren15,loren16,lamb16}  

%Understanding dusty gas dynamics is
%therefore essential to any planet formation theory. 
%phenomena that depends on gas drag 
%planet formation in dusty disks 
%dust component is building block of planets
%understanding dusty gas dynamics 
%small amount of dust can have significant consequences 

%particles 
Current state-of-the-art models of dusty protoplanetary
disks directly simulate gas dynamics with explicit
Lagragian dust particles \citep{nelson10,bai10,johansen11,yang14,zhu14,gibbons15,simon16,baruteau16}. In this
approach, the equations of motion for each particle is solved
including dust-gas drag, the strength of which is measured 
by a stopping time $\tstop$ --- the decay timescale for the 
relative velocity between gas and dust. Thus small $\tstop$
corresponds to tightly-coupled particles. 

%refs 

Another approach is to model the dust population as a continous, pressureless
fluid
\citep{paardekooper06b,meheut12,laibe12,loren14,fu14b,surville16}. The equations
hydrodynamics are solved for two fluids: the gas and dust with density and
velocity $\rhog$, $\bm{v}_\mathrm{g}$ and $\rhod$,
$\bm{v}_\mathrm{d}$, respectively. Source terms are introduced into
the momentum/energy equations to model dust-gas drag. This approach
can take advantage of, and adapt, already-developed methods/codes for
simulating pure gas dynamics. 

%refs 

It is possible to re-formulate the two-fluid equations into that for a
single fluid of for the total density $\rho$ and center-of-mass
velocity $\bm{v}$ of the dusty gas mixture
\citep{laibe14}. The standard hydrodynamic equations for
$\rho$ and $\bm{v}$ are supplemented by evolutionary equations for
the dust-to-gas ratio $\rhod/\rhog$ and the dust-gas velocity
difference $\Delta\bm{v}$. The latter also contributes to
additional source terms in the momentum and energy equations.  
The one-fluid approach is appropriate for 
simulating gas tightly coupled to dust particles with small $\tstop$.  
This can be numerically challenging for the two-fluid or Lagragian
particles approach; but is straight forward to model in the
one-fluid framework \citep{price15}.  

The single fluid formalism also has the important advantage that it is
amenable to standard analytical methods
\citep{youdin05a,jacquet11}. This is the necessary  approach to gain physical insight.  
In this work, we are motivated to develop a general analytical
framework to  study the stability of dusty protoplanetary disks by
drawing on the physical and mathematical parallel between dusty gas and standard
hydrodynamics. This is possible in certain limits of astrophysical
interest.  

We consider tightly coupled dust particles with $\tstop\Omega\ll
1$, where $\Omega$ is the disk orbital frequency. We may then use the
`terminal velocity approximation' to set $\Delta \bm{v} = \tstop\nabla
P/\rhog$, where $P$ is the gas pressure \citep{youdin05a,
  jacquet11,laibe14}. We also consider irradiated protoplanetary disks 
in which the temperature is set externally \citep{chiang97,stam08}. We
may then adopt a locally isothermal equation of state $P = c_s^2\rhog$
where $c_s$ is a prescribed sound-speed profile.  This physically
represents short cooling times \citep{lin15}. 

Under these approximations, we show that the one-fluid equations of 
isothermal dusty gas dynamics are equivalent to that of adiabatic, pure gas
dynamics. This is because tightly-coupled dust particles are almost
entrained in the gas; and this property is analogous to the advection
of entropy following an adiabatic fluid. We will formally relate the
entropy of isothermal dusty gas to its dust content. Finite 
dust-gas drag can be re-interpreted as heating/cooling: the relative
motion between dust and gas causes a fluid parcel to exchange dust
particles with its surroundings; an effect similar to heat exchange
between a gas parcel and its surroundings.   

%suitable for 
%analytic studies 
%more importantly, can understood 

By re-casting the dusty problem into identical form to 
hydrodynamics, we can apply standard methods to generalize previous stability 
results for pure gas disks to the dusty case. We demonstrate this for
perfectly coupled dust in strictly and locally isothermal disks. We
show that sharp dust edges may destabilize protoplanetary disks.  
We also study the effect of dust-loading on the vertical shear
instability \citep[VSI,][]{nelson13,lin15,barker15}. We provide 
thermodynamical interpretations of our dusty stability calculations. 

This paper is organized as follows. In \S\ref{setup} we list the
equations describing tightly-coupled dusty gas. We show they are in 
fact the usual hydrodynamic equations. This leads to a precise 
definition of the entropy of an isothermal dusty gas. We describe approximate 
equilibrium solutions in \S\ref{eqm}. In \S\ref{limits} we discuss
general stability properties of the dusty problem in various
limits. We use these results to constraint the sharpness of dust edges
in isothermal disks. In \S\ref{linear_problem} we formulate an
explicit stability calcultion to study the dusty VSI, for which we
present numerical results in \S\ref{results}. We 
discuss future applications of the dusty/adiabatic gas equivalence in 
\S\ref{discussion} before summarizing in \S\ref{summary}. 

\input{model}
\input{linear}
\input{result}
\input{summary}

\appendix
\input{appendix}

\bibliographystyle{apj}
\bibliography{ref}


\end{document}

